Probabilistic inference in dynamical systems

### 2024-04-10 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): add inference model description Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

If we are just given samples of pre-mRNA and mRNA counts without additional
information about their sampling times, we can describe the data set 
$\mathcal{D}$ consisting of pre-mRNA and mRNA counts for a number 
of genes $G$ across a number of cells $N$

$$
\mathcal{D} = \left\{ (u_{ij}, s_{ij}) \mid i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\}
$$

@@

- $u_{ij}$ represents the count of pre-mRNA for gene $i$ in cell $j$,
- $s_{ij}$ represents the count of mRNA for the same gene $i$ in the same cell $j$,
- $G$ is the total number of genes in the study,
- $N$ is the total number of cells sampled.

@@

1. **Normalization**: Transforming the observed counts $u_{ij}$ and $s_{ij}$ to 
   their dimensionless counterparts $u^{\ast}_{ij}$ and $s^{\ast}_{ij}$ using 
   any reasonable concentration scale $U_0$.
2. **Model Fitting**: Applying statistical methods to estimate the latent variables 
   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$ that best 
   fit the observed data.
3. **Evaluation**: Assessing the fit of the model and the estimated parameters'
   biological plausibility and consistency across different genes and cells.

This representation of the dataset and the associated objective of 
statistical learning provides a solid foundation for applying probabilistic
modeling techniques to calibrate an inference procedure andevaluate the 
plausibility of this model in such a manner that we can eventually compare 
multiple candidate models to one another.

@@

Now that we have illustrated how to simulate the system, we can proceed to
define a probabilistic model and perform inference on its latent variables.

>>>>

If we are given samples of pre-mRNA and mRNA counts along with
the sampling time of each cell, we can describe the data set 
$\mathcal{D}$ consisting of pre-mRNA and mRNA counts for a number 
of genes $G$ across a number of cells $N$

\begin{align}
\mathcal{D} = \left\{ (t_j, u_{ij}, s_{ij}) \mid 
  i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\} \label{eq-dataset},
\end{align}

@@

- $t_j$ represents the sampling time for cell $j$,
- $u_{ij}$ represents the count of pre-mRNA for gene $i$ in cell $j$,
- $s_{ij}$ represents the count of mRNA for the same gene $i$ in the same cell $j$,
- $G$ is the total number of genes in the study,
- $N$ is the total number of cells sampled.

@@

1. **Normalization**: Transforming the observed counts $u_{ij}$ and $s_{ij}$ to 
   their dimensionless counterparts $u^{\ast}_{ij}$ and $s^{\ast}_{ij}$ using 
   a reasonable concentration scale $U_0$.
2. **Model Fitting**: Applying statistical methods to estimate the latent 
   variables 
   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0 \right)$ 
   that best fit the observed data.
3. **Evaluation**: Assessing the fit of the model and the estimated parameters'
   biological plausibility and consistency across different genes and cells.

We will return later to the alternative version of this problem where $t_j$ 
are not observed and must be inferred along with the latent variables 
$\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$.

This representation of the dataset and the associated objective of 
statistical learning provides a solid foundation for applying probabilistic
modeling techniques to calibrate an inference procedure and evaluate the 
plausibility of this model in such a manner that we can eventually compare 
multiple candidate models to one another.

@@

Now that we have illustrated how to simulate a prototypical system, we can
proceed to define a probabilistic model and perform inference on its latent
variables given data like \eqref{eq-dataset}. In general we should consider multiple
models with different assumptions and levels of complexity, but for now we will
focus on one of the simpler models to illustrate the process. In particular, in
most cases, when performing inference on the latent variables of a dynamical
system, we assume that the observation time points are fixed or otherwise
observed. We will begin with this assumption and then proceed to discuss how to
relax it.

In the case where we can indeed estimate analytical solutions to the system, we
can consider a model with the structure proposed in 
(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) where, given a number of 
genes $G$ and cells $N$, for each
$(i,j) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\}$ we have

\begin{align}
    u^{\ast}_{0i}, s^{\ast}_{0i} &\sim 
      \text{LogNormal}(\mu_{0}, \sigma_{0}^2), 
      \label{eq-init-conds-priors-gene} \\

    \gamma^{\ast}_i &\sim 
      \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2),
      \label{eq-gamma-prior-gene} \\

    \sigma_{ui}, \sigma_{si} &\sim 
      \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2),
      \label{eq-noise-std-priors-gene} \\

    u^{\ast}_{ij} &= 
      1 + (u^{\ast}_{0i} - 1) \cdot e^{-t^{\ast}_j},
      \label{eq-u-star-model-cell} \\

    s^{\ast}_{ij} &= 
      \begin{cases}
        \frac{1}{\gamma^{\ast}_i} + 
        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i t^{\ast}_j} + 
        \xi_i \cdot e^{-t^{\ast}_j},& \gamma^{\ast}_i \neq 1 \\

        1 + (s^{\ast}_{0i} - 1) e^{-t^{\ast}_j} + 
          (u^{\ast}_{0i} - 1) t^{\ast}_j e^{-t^{\ast}_j},& \gamma^{\ast}_i = 1 \\
      \end{cases}, \label{eq-s-star-model-cell} \\

    \hat{u}^{\ast}_{ij} &\sim \text{Normal}(u^{\ast}_{ij}, \sigma_{ui}^2), 
    \label{eq-u-obs-cell} \\

    \hat{s}^{\ast}_{ij} &\sim \text{Normal}(s^{\ast}_{ij}, \sigma_{si}^2). 
    \label{eq-s-obs-cell}
\end{align}

The model description in (\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) 
can be associated to the plate diagram in 
@fig-transcription-splicing-degradation-graphical-model-1.

```{python}
# | label: fig-transcription-splicing-degradation-graphical-model-1
# | code-fold: true
# | fig-cap: Graphical model for the transcription-splicing-degradation model.
import daft
import matplotlib.pyplot as mpl

mpl.rcParams["font.family"] = "serif"
mpl.rcParams["font.size"] = 16
mpl.rcParams["text.usetex"] = True

pgm = daft.PGM(line_width=1.2)

# hyperparameters
pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
pgm.add_node("mu_gamma", r"$\mu_{\gamma}$", 2.5, 6, fixed=True)
pgm.add_node("sigma_gamma", r"$\sigma_{\gamma}^2$", 3.5, 6, fixed=True)
pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)

# latent variables for gene-specific parameters
pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)

# latent variables for cell-specific outcomes
pgm.add_node(
    "u_star_ij",
    r"$u^{\ast}_{ij}$",
    2,
    4,
    scale=1.0,
    shape="rectangle",
)
pgm.add_node(
    "s_star_ij",
    r"$s^{\ast}_{ij}$",
    4,
    4,
    scale=1.0,
    shape="rectangle",
)

# observed data
pgm.add_node(
    "t_star_j",
    r"$t^{\ast}_j$",
    6.0,
    3.25,
    observed=True,
    shape="rectangle",
)
pgm.add_node(
    "u_obs_ij",
    r"$\hat{u}^{\ast}_{ij}$",
    2,
    2.5,
    scale=1.0,
    observed=True,
)
pgm.add_node(
    "s_obs_ij",
    r"$\hat{s}^{\ast}_{ij}$",
    4,
    2.5,
    scale=1.0,
    observed=True,
)

# edges
edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)

pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)

pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)

pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)

# plates
pgm.add_plate([0.5, 1.2, 5, 4.4], label=r"$i \in \{1, \ldots, G\}$", shift=-0.1)
pgm.add_plate(
    [1.0, 1.8, 5.5, 2.75], label=r"$j \in \{1, \ldots, N\}$", shift=-0.1
)

pgm.render()
```

Of course, in general, we will not have access to the analytical solutions, so
we will need to simulate the system and then perform inference on the latent
variables as suggested in (\ref{eq-init-conds-priors-num}--\ref{eq-s-obs-num}).

\begin{align}
    u^{\ast}_0, s^{\ast}_0 &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \label{eq-init-conds-priors-num}, \\
    \gamma^{\ast} &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-num}, \\
    \sigma_u, \sigma_s &\sim \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2) \label{eq-noise-std-priors-num}, \\
    (u^{\ast}, s^{\ast}) &= \text{NumericalSolver}\left(\frac{du^{\ast}}{dt^{\ast}}, \frac{ds^{\ast}}{dt^{\ast}}, u^{\ast}_0, s^{\ast}_0, \gamma^{\ast}, t^{\ast}\right) \label{eq-numerical-solution}, \\
    \hat{u}^{\ast} &\sim \text{Normal}(u^{\ast}, \sigma_u^2) \label{eq-u-obs-num}, \\
    \hat{s}^{\ast} &\sim \text{Normal}(s^{\ast}, \sigma_s^2) \label{eq-s-obs-num}.
\end{align}

Where $\frac{du^{\ast}}{dt^{\ast}}$ and $\frac{ds^{\ast}}{dt^{\ast}}$, 
are given by (\ref{eq-dustardtstar}-\ref{eq-dsstardtstar}) and noting that 
we suppress the indices from 
(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) for brevity.
